---
title: "Breast_cancer"
output:
  pdf_document: default
  html_document: default
date: "2022-10-24"
---

# question
Find the best treatment corresponding of the patient situation. ..à  reformuler  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
library(tidyverse)
library(survival)
library(survminer)
```

# Load the dataset and data cleaning

```{r include=FALSE}
# load the dataset 
df <- read.csv("METABRIC_RNA_Mutation.csv")

# Keep the first 31 clinical features
df <- df[0:31]

# Keep the feature of interest
features <- c("patient_id", "age_at_diagnosis", "type_of_breast_surgery", "cancer_type_detailed", "chemotherapy", "hormone_therapy", "primary_tumor_laterality", "overall_survival_months", "radio_therapy","tumor_stage", "death_from_cancer" )

# Create a dataframe, df_subset for the anlysis
df_subset <- df[,features]

#The type_of_breast_surgery variables should contains only 'BREAST CONSERVING','MASTECTOMY' value, 
#but we can observe there are 22 blank value (not exactly NA). It is not usable, so will remove them
df_subset <- df_subset[df_subset$type_of_breast_surgery %in% c('BREAST CONSERVING','MASTECTOMY'),]
# same is true for primary_tumor_laterality (11 blank value)
df_subset <- df_subset[df_subset$primary_tumor_laterality %in% c('Left','Right'),]
df_subset <- drop_na(df_subset)

head(df_subset)
```


```{r}

```


For our analysis we have a dataframe with the status "death_from_cancer", the survival time "overall_survival_months" and 9 other features.

Categorical features:

-   hormone_therapy: 0 no therapy, 1 with therapy
-   primary_tumor_laterality: right/left indicates one which side the cancer appeared
-   radio_therapy: 0 no therapy, 1 with therapy
-   tumor_stage: from 0 to 4
-   chemotherapy: 0 no therapy, 1 with therapy
-   type_of_breast_surgery: MASTECTOMY fro full ablation, BREAST CONSERVING otherwise
-   cancer_type_detailed: Breast Invasive Ductal Carcinoma or Breast Mixed Ductal and Lobular Carcinoma, it gives the localisation of the tumor
-   patient_id

Continuous features:

-   age_at_diagnosis
-   overall_survival_months

There is no null value for the survival in months variable and the distribution indicates people roughly survive between 0 to 400 months after the breast surgery.



## Convention for the status

For our categorical features we my need to specify explicitly the level for better explainability.

### Death from cancer convention

We want to change the columns

-   0: living or died from other causes
-   1: died because of the cancer

```{r include=FALSE}
df_subset <- df_subset %>%
  rename(status = "death_from_cancer",
         duration= "overall_survival_months")%>%
  mutate(
    status = ifelse(status == "Died of Disease", 1, 0)
  )
  
  
```

### Type of surgery convention

We want to level the variable type of breast surgery, will put breast conserving as the reference level (0)

```{r include=FALSE}
df_subset <- df_subset |> 
  #mutate(type_of_breast_surgery  = ifelse(tumor_stage %in% c(0,4), 'other', tumor_stage)) |> 
  mutate(type_of_breast_surgery  = factor(type_of_breast_surgery )) |>
  mutate(type_of_breast_surgery = relevel(type_of_breast_surgery , ref = 'BREAST CONSERVING'))
table(df_subset$type_of_breast_surgery)
```

# Overall non parametric survival analysis

Let's have a look on the survival time of the patients.

```{r}
## Nelson-AAlen estimator
fit_na <- survfit(Surv(duration, status) ~ 1, data = df_subset, type = "fh")
plot(fit_na, 
     main ="Nelson Aalen survival estimator",
     xlab = "Months",
     ylab= "Survival function")
```

```{r include=FALSE}
fit_na
```

We notice from this survival analysis there are 50% of survivals after 286 months.



## Kaplan-meier with one feature and 2 groups: chemiotherapy

Now we want to analyse the survival 

```{r}
surv_chem <- survfit(Surv(duration, status ) ~ chemotherapy, data = df_subset)
plot(surv_chem, col= (1:2), main ="Kaplan-Meier", xlab = "Months", ylab = "Survival function")
legend("topright", legend= names(surv_chem$strata),
       col=c("black", "red"), lty=1)
#summary(surv_chem)
```
```{r}
# Change color, linetype by strata, risk.table color by strata
ggsurvplot(surv_chem,
          pval = TRUE, conf.int = TRUE,
          xlab = "Time in months",
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          ncensor.plot = TRUE,
          palette = c("#E7B800", "#2E9FDF"))
```



```{r}
# Summary from the survminer package
res.sum <- surv_summary(surv_chem)
attr(res.sum, "table")
```
> We can observe it seems there is a difference between the 2 groups. indeed, people without any chemiotherapy tend to survive longer  50 % of survivals after 301 months against 147 for people with chemiotherapy.
Let's have a look on the significance of this difference with a LogRank.

## LogRank

Let's define the null hypothesis H0 : the two groups have the same survival distributions.

```{r}
logrank_chem <- survdiff(Surv(duration, status ) ~ chemotherapy, data = df_subset)
logrank_chem
```

> The p-value is << before 0.05. So we can reject the null hypothesis and accept there is a significance difference between the two groups.

## Kaplan Meir analysis for all the binary  variables

> Treatments
  * type_of_breast_surgery
  * chemotherapy
  * hormone_therapy
  * radio_therapy
  
> Breast side affected: primary_tumor_laterality

```{r}
covariates <- c("type_of_breast_surgery" , "chemotherapy" , "hormone_therapy" , "radio_therapy", "primary_tumor_laterality")

univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(duration, status)~', x)))
                        
univ_km <- lapply(univ_formulas, 
                   function(x){
                     logrank_univ <- survdiff(x, data = df_subset)
                     }
                   )

univ_km_pvalue <- lapply(univ_km,
                         function(x){
                           signif(1 - pchisq(x$chisq, length(x$n) - 1), 2)
                         }
                        )

km_table <- t(as.data.frame(univ_km_pvalue, check.names = FALSE))
colnames(km_table)<-"p.value"

as.data.frame(km_table)
```
> Looking at all binary variables, we can observe that only the "type_of_breast_surgery" and the "chemotherapy" have a significan impact on the survival.

## Kaplan Meir fit with chemotherapy and radio therapy

```{r}
multiple_fit_km <- survfit(Surv(duration, status ) ~ chemotherapy + radio_therapy, data = df_subset)
multiple_fit_km
```

```{r}
plot(multiple_fit_km, col= (1:4), main ="Kaplan-Meier", xlab = "Months", ylab = "Survival functions")
legend("topright", legend= names(multiple_fit_km$strata),
       col=c(1:4), lty=1 )
```
We can observe in this additive model 4 groups:
*   without chemotherapy and without radio therapy: black line
*   without chemotherapy and with radio therapy: red
*   with chemotherapy and without radio therapy: green
*   with chemotherapy and with radio therapy: blue


From this plot we may have a doubt on the significance difference between the group in red (without chemotherapy and with radio therapy) and the one in black ( without chemotherapy and without radio therapy) as there are close together.
Let's do a log rank test.

```{r}
logrank_multiple <- survdiff(Surv(duration, status ) ~ chemotherapy + strata(radio_therapy), data = df_subset)
```

```{r}
logrank_multiple
```

```{r}
#survdiff(Surv(duration, status ) ~ chemotherapy[0] + radio_therapy[0], data = df_subset)

```
# Conclusion for categorical features


# Parametric survival analysis


## Defining the best treatement for a designed group of person

The overall goal is to provide recommendation on the most efficient treatment (higher survival in months) for a designed group of person. The design group of person will be done with the use of the features:

1)
-   age_at_diagnosis
-   primary_tumor_laterality
-   tumor stage
-   cancer_type_detailed

2)
For the recommendation , we will use the followings treatments:
 -    type_of_breast_surgery
 -    chemotherapy
 -    hormone_therapy
 -    radio_therapy
 
 
### Defining the groups of patients

To define the groups, we will apply a Cox analysis, and check which variables has a significant impact on the survival.

```{r}
cox_group <- coxph(Surv(duration, status) ~ age_at_diagnosis + primary_tumor_laterality +  tumor_stage + cancer_type_detailed,  data = df_subset)

summary(cox_group)
```

It seems the age as continuous variable is not significant on the survival in months (p value = 0.97); Let's check if we change the step for the age_at_diagnosis value from 1 year to 10.


> Let's check first quickly the distribution of the age. The feature is continuous.

```{r}
hist(df_subset$age_at_diagnosis)
```

```{r}
cox_group_age <- coxph(Surv(duration, status) ~ I(age_at_diagnosis/10) + primary_tumor_laterality +  tumor_stage + cancer_type_detailed,  data = df_subset)

summary(cox_group_age)
```

By changing the step of the age variable from one year to ten do not have a  significant impact (p value = 0.97)

Just to confirm, we will have a look on a model using only the Age/10 variable.

```{r}
coxph(Surv(duration, status) ~ I(age_at_diagnosis/10) ,  data = df_subset) |>
  summary()
```
> It is a bit more significant when using alone (p value = 0.397), but it is far from being significant on the survivals of the patients. We wil remove it from the model.


> Conclusion for the designed group.

It seems in our additive model (age_at_diagnosis + primary_tumor_laterality +  tumor_stage + cancer_type_detailed) only the tumor stage has an impact on the overall survival. We will use this one to discriminate the groups and select the best treatment, or combination of treatment accordingly the tumor stage.

Let's compute a cox model using only the tumor stage variable.



### Cox Analys with one features and multiple groups: 


We will look at the tumor stage variable:

```{r}
boxplot(df_subset$duration ~ df_subset$tumor_stage,xlab='Tumor Stage', ylab='Survival time (months)', main='Survival by Tumor Stage')
```

> We notice there is a trend indicating the more the tumor stage is advanced, the higher the risk for people to die from breast cancer (less months to survive). Let's do a Cox analysis to confirm this hypothesis.

Let's review the number of elements per groups
```{r}
table(df_subset$tumor_stage) # show number of elements per group
```
Let's first define the reference level.
Stage0 being a small group (4) there is a chance that it will impact the significance of the comparaison with the other groups.
This is confirmed by the following analysis where the pvalue of all groups is high (>>0.05).

```{r}
df_subset <- df_subset |> 
  #mutate(tumor_stage = ifelse(tumor_stage %in% c(0,4), 'other', tumor_stage)) |> 
  mutate(tumor_stage = factor(tumor_stage)) |>
  mutate(tumor_stage = relevel(tumor_stage, ref = '0'))
level.cph <- coxph(Surv(duration, status) ~ tumor_stage, data = df_subset)
summary(level.cph)
```

It will make more sense to choose Stage1 as a reference.


```{r}
df_subset <- df_subset |> 
  mutate(tumor_stage = factor(tumor_stage)) |>
  mutate(tumor_stage = relevel(tumor_stage, ref = '1')) # explicitly define stage1 as the reference
levels(df_subset$tumor_stage) # display the level
```
Let's analyse the tumor_stage compare to group '1'

```{r}
level.cph <- coxph(Surv(duration, status) ~ tumor_stage, data = df_subset)
summary(level.cph)
```
We can observe:
  * Stage0 has a high pvalue, so we cannot ignore the null hypothesis for this group.
  * Stage2, Stage3 and Stage4 has a small pvalue (<0.05), so their difference is significant.
  * for Stage2, Stage3 and Stage4 $\beta$ is positive so there a higher risk for these groups compare to Stage1 group
  * There is 2.0 [1.6, 2.5] more risk for Stage2, 4.5[3.3, 6.2] for Stage3 and 8.5[4.0, 18.3] for Stage4 to die compare to Stage1.
  * The confidence interval of Stage4 is higher than for Stage2 and Stage3, possibly due to the low number of elements in group.



## Defining the best treatment according to the tumor stage


### Defining the best treatment for patient having a tumor stage 1

By best treatment, we mean a treatment the give the longer survival in months.

Let's compute our first cox model using all the treatment in an additive model then conduct a variable selection.

```{r}
cox_group_1 <- coxph(Surv(duration, status) ~ type_of_breast_surgery + chemotherapy + hormone_therapy + radio_therapy, data = df_subset[df_subset$tumor_stage==1,])
```

```{r}
summary(cox_group_1)
```
> We notice there is only the chemotherapy that has a significant impact (p = 0.0002). Surprisingly, from the analysis we can infer the risk to die from cancer breast is 4 times higher for the patients. 


### Defining the best treatment for patient having a tumor stage 2

By best treatment, we mean a treatment the give the longer survival in months.

Le'ts compute our first cox model using all the treatment in an additive model then conduct a variable selection.

```{r}
cox_group_2 <- coxph(Surv(duration, status) ~ type_of_breast_surgery + chemotherapy + hormone_therapy + radio_therapy, data = df_subset[df_subset$tumor_stage==2,])

```

```{r}
summary(cox_group_2)
```
Both the type of breast surgery and chemotherapy have a significant impact p value respectively 0.04 and 0.03). For both of them the risk is increased by 35 %. Hormone therapy is not really far from being significant. Let's create a model with a variable selection with the step AIC algorithm.

Let's see if we can apply a varibale selection on this model.

```{r}
step(cox_group_2)
```
> This time only the type of breast surgery is significant (p = 0.004) but has an negative impact on the survival in months. However The chemotherapy (p value 0.058) and hormone therapy (p value 0.053) are closed from being significant and we can notice this time the hormone therapy decreased the risk to die exp(beta)=0.77<1

We will keep the final model with only type of breast surgery

```{r}
cox_group_2 <- coxph(Surv(duration, status) ~ type_of_breast_surgery , data = df_subset[df_subset$tumor_stage==2,])
cox_group_2
```
> Our final model for the group 2 is defined onnly by the treatment of the breast surgery which has a significant impact (p = 0.004) and increased the risk by 48%.

### Defining the best treatment for patient having a tumor stage 3

By best treatment, we mean a treatment the give the longer survival in months.

Le'ts compute our first cox model using all the treatment in an additive model then conduct a variable selection.

```{r}
cox_group_3 <- coxph(Surv(duration, status) ~ type_of_breast_surgery + chemotherapy + hormone_therapy + radio_therapy, data = df_subset[df_subset$tumor_stage==3,])

```

```{r}
summary(cox_group_3)
```
> the type of surgery is closed from being significant (p =0.059) and the hotmone therapy is isgnificant (p=0.019) and decreased the risk beta is 0.29.

```{r}
covariates <- c("type_of_breast_surgery" , "chemotherapy" , "hormone_therapy" , "radio_therapy")
univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(duration, status)~', x)))
                        
univ_models <- lapply( univ_formulas, function(x){coxph(x, data = df_subset[df_subset$tumor_stage==3,])})
# Extract data 
univ_results <- lapply(univ_models,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (", 
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                        "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })
res <- t(as.data.frame(univ_results, check.names = FALSE))
as.data.frame(res)
```


### Defining the best treatment for patient having a tumor stage 4

By best treatment, we mean a treatment the give the longer survival in months.

Le'ts compute our first cox model using all the treatment in an additive model then conduct a variable selection.

```{r}
cox_group_4 <- coxph(Surv(duration, status) ~ type_of_breast_surgery + chemotherapy + hormone_therapy + radio_therapy, data = df_subset[df_subset$tumor_stage==4,])

```

```{r}
summary(cox_group_4)
```
> Unfortunately, there is no significant treatment for patient at stage 4. We can have a look on the models with univariate cox model for each treatment alone.


```{r}
treatments <- c("type_of_breast_surgery" , "chemotherapy" , "hormone_therapy" , "radio_therapy")
univ_formulas <- sapply(treatments,
                        function(x) as.formula(paste('Surv(duration, status)~', x)))
                        
univ_models <- lapply( univ_formulas, function(x){coxph(x, data = df_subset[df_subset$tumor_stage==4,])})
# Extract data 
univ_results <- lapply(univ_models,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (", 
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                        "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })
res <- t(as.data.frame(univ_results, check.names = FALSE))
as.data.frame(res)
```

> Same conclusion. No significant treatment.

# Cox model assumptions

We will check assumptions on the final model for patients at tumor stage 2.

```{r}
cox_group_2
```

## Testing proportional hazard

### Check the proportional hazard with a kaplan meier 

We will plot a log-log plot and check if the two survival curves are parallels.



```{r}
km_breast_surgery <- survfit(Surv(duration, status) ~ type_of_breast_surgery, data = df_subset[df_subset$tumor_stage == 2, ])
ggsurvplot(km_breast_surgery, fun = "cloglog")
```


> A first look gives us good chance to have proportional hazards. Let's have a look on the Schoenfeld residuals over time.


```{r}
test.ph <- cox.zph(cox_group_2)
ggcoxzph(test.ph)
```
We can see we are centered around zero and all residuals are between $+/- 2\sigma$ and the p-valueis 0.79 (not significant) for this test which indicates we can accept H0, that is there is a proportional hazard.

### Conclusion
We can accept the validity of the assumptions for the cox model to define the treatment by breast surgery for the group of patients defined by having a tumor stage 2.


# Conclusion



We wanted to propose the best combination of treatments depending on characteristics of people (age, tumor stage, side of the primary tumor and type of cancer).

First we have check the impact of the having a chemotherapy on all patients regardless any other features. It shows at the first sight it has a significant and negative influence on the survival time as half of the patient with chemotherapy were died after 163 months against 300 for those without chemotherapy.

It is surprising, but we have to be careful it is only a correlation and not a causality.

Second We have  define which of these characteristics were the most significant for the survival time and we have kept only the tumor stage which were highly significant (p value $2e-16$). 

Thanks to this analysis, we have designed four groups depending on their tumor stage.

Then we have checked for the best combination of treatments for each of these groups by starting from additive models and applying variables selections with a step wise approach by the the Akaike criterion.

Globally we have observed patients following a chemotherapy for all groups is significant and has a negative impact i.e. it decreases the survival time. Only the hormone therapy seems to have a significant(p-value 0.0019 ) and positive impact (decrease the risk by 3) for patients having a tumor stage 3.

From this conclusion, we may infer that it should be interesting to check these treatments for more detailed groups with more characteristics discriminating them than only the tumor stage, otherwise the chemotherapy would have no reasons to exist.

















