---
title: "Breast_cancer"
output: html_document
date: "2022-10-24"
---

# question
Find the best treatment corresponding of the patient situation. ..à  reformuler  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load library
```{r}
library(tidyverse)
library(survival)
```

# Load the dataset
```{r}
df <- read.csv("METABRIC_RNA_Mutation.csv")
```
## Context 



## Subset the dataset
```{r}
# Keep the first 31 clinical features
df <- df[0:31]
str(df)
```
## Keep the feature of interest

```{r}
features <- c("patient_id", "age_at_diagnosis", "type_of_breast_surgery", "cancer_type_detailed", "chemotherapy", "hormone_therapy", "primary_tumor_laterality", "overall_survival_months", "radio_therapy","tumor_stage", "death_from_cancer" )
```

```{r}
df_subset <- df[,features]
str(df_subset)
```
## Review duration values

```{r}
# Check if there is NA
sum(is.na(df_subset$overall_survival_months))

# plot the survival
hist(df_subset$overall_survival_months)
```


## remove empty and NA values

```{r}
table(df_subset$type_of_breast_surgery)

df_subset <- df_subset[df_subset$type_of_breast_surgery %in% c('BREAST CONSERVING','MASTECTOMY'),]
df_subset <- drop_na(df_subset)

```


```{r}
summary(df_subset)
```

# Convention for the status

We want to change the columns
0: living or died from other causes
1: died because of the cancer

```{r}
df_subset <- df_subset %>%
  rename(status = "death_from_cancer",
         duration= "overall_survival_months")%>%
  mutate(
    status = ifelse(status == "Died of Disease", 1, 0)
  )
  
  
```

# Overall non parametric survival analysis

```{r}
## Nelson-AAlen estimator
fit_na <- survfit(Surv(duration, status) ~ 1, data = df_subset, type = "fh")
plot(fit_na, 
     main ="Nelson Aalen survival estimator",
     xlab = "Months",
     ylab= "Survival function")
```

```{r}
fit_na
```

> We notice from this survival analysis there are 50% of surivals after 286 months.




## Kaplan-meier with one features and 2 groups: chemiotherapy

```{r}
surv_chem <- survfit(Surv(duration, status ) ~ chemotherapy, data = df_subset)
plot(surv_chem, col= (1:2), main ="Kaplan-Meier", xlab = "Months", ylab = "Survival function")
legend("topright", legend= names(surv_chem$strata),
       col=c("black", "red"), lty=1)
#summary(surv_chem)
```

```{r}
surv_chem
```
> We can observe it seems there is a difference between the 2 groups. indeed, people without any chemiotherapy tend to survive longer  50 % of survivals after 301 months against 147 for people with chemiotherapy.
Let's have a look on the significance of this difference with a LogRank.

## LogRank

Let's define the null hypothesis H0 : the two groups have the same survival distributions.

```{r}
logrank_chem <- survdiff(Surv(duration, status ) ~ chemotherapy, data = df_subset)
logrank_chem
```

> The p-value is << before 0.05. So we can reject the null hypothesis and accept there is a significance difference between the two groups.
Now we will look at an additibve model by includeing with the chemotherapy the raio therapy.


## Kaplan Meir fit with chemotherapy and radio therapy

```{r}
multiple_fit_km <- survfit(Surv(duration, status ) ~ chemotherapy + radio_therapy, data = df_subset)
multiple_fit_km
```

```{r}
plot(multiple_fit_km, col= (1:4), main ="Kaplan-Meier", xlab = "Months", ylab = "Survival functions")
legend("topright", legend= names(multiple_fit_km$strata),
       col=c(1:4), lty=1 )
```
We can observe in this additive model 4 groups:
*   without chemotherapy and without radio therapy: black line
*   without chemotherapy and with radio therapy: red
*   with chemotherapy and without radio therapy: green
*   with chemotherapy and with radio therapy: blue


From this plot we may have a doubt on the significance difference between the group in red (without chemotherapy and with radio therapy) and the one in black ( without chemotherapy and without radio therapy) as there are close together.
Let's do a log rank test.

```{r}
logrank_multiple <- survdiff(Surv(duration, status ) ~ chemotherapy + strata(radio_therapy), data = df_subset)
```

```{r}
logrank_multiple
```

```{r}
#survdiff(Surv(duration, status ) ~ chemotherapy[0] + radio_therapy[0], data = df_subset)
```
# Conclusion for categorical features

# Parametric survival analysis

## Cox Analys with one features and multiple groups: 

Let's review the dataset
```{r}
head(df_subset)
```
We will look at the tumor stage variable:

```{r}
boxplot(df_subset$duration ~ df_subset$tumor_stage,xlab='Tumor Stage', ylab='Survival time (months)', main='Survival by Tumor Stage')
```

Let's review the number of elements per groups
```{r}
table(df_subset$tumor_stage) # show number of elements per group
```
Let's first define the reference level.
Stage0 being a small group (4) there is a chance that it will impact the significance of the comparaison with the other groups.
This is confirmed by the following analysis where the pvalue of all groups is high (>>0.05).

```{r}
df_subset_relevelled <- df_subset |> 
  #mutate(tumor_stage = ifelse(tumor_stage %in% c(0,4), 'other', tumor_stage)) |> 
  mutate(tumor_stage = factor(tumor_stage)) |>
  mutate(tumor_stage = relevel(tumor_stage, ref = '0'))

level.cph <- coxph(Surv(duration, status) ~ tumor_stage, data = df_subset_relevelled)
summary(level.cph)
```

It will make more sense to choose Stage1 as a reference.


```{r}
df_subset <- df_subset |> 
  mutate(tumor_stage = factor(tumor_stage)) |>
  mutate(tumor_stage = relevel(tumor_stage, ref = '1')) # explicitly define stage1 as the reference

levels(df_subset_relevelled$tumor_stage) # display the level
```
Let's analyse the tumor_stage compare to group '1'

```{r}
level.cph <- coxph(Surv(duration, status) ~ tumor_stage, data = df_subset_relevelled)
summary(level.cph)
```
We can observe:
  * Stage0 has a high pvalue, so we cannot ignore the null hypothesis for this group.
  * Stage2, Stage3 and Stage4 has a small pvalue (<0.05), so their difference is significant.
  * for Stage2, Stage3 and Stage4 $\beta$ is positive so there a higher risk for these groups compare to Stage1 group
  * There is 2.3[1.6, 2.5] more risk for Stage2, 4.6[3.4, 6.3] for Stage3 and 9.3[4.5, 19.2] for Stage4 to die compare to Stage1.
  * The confidence interval of Stage4 is higher than for Stage2 and Stage3, possibly due to the low number of elements in group.

```{r}
table(df_subset$cancer_type)
table(df_subset$type_of_breast_surgery)
```
```{r}
summary(df_subset_relevelled)
```

# Misc. Notes
We want to level the variable type of breast surgery, will put breast conservering as the reference level (0)

```{r}
df_subset <- df_subset |> 
  #mutate(type_of_breast_surgery  = ifelse(tumor_stage %in% c(0,4), 'other', tumor_stage)) |> 
  mutate(type_of_breast_surgery  = factor(type_of_breast_surgery )) |>
  mutate(type_of_breast_surgery = relevel(type_of_breast_surgery , ref = 'BREAST CONSERVING'))
table(df_subset$type_of_breast_surgery)
```

type_of_breast_surgery


```{r}

coxph(Surv(duration, status) ~ type_of_breast_surgery + chemotherapy + hormone_therapy + radio_therapy, data = df_subset) |> summary()
```
# Variable selection
In this model, chemotherapy and type of breast surgery mastectomy has p value under 0.05 => significant impact. It increases the risk of dying respectively by 1.6 and 1.8.
We need to be careful to infer causality we will have to consider other variables.

```{r}
coxph(Surv(duration, status) ~ tumor_stage + cancer_type_detailed+ primary_tumor_laterality , data = df_subset) |> summary()

```
```{r}
coxph(Surv(duration, status) ~ cancer_type_detailed, data = df_subset) |> summary()
```
