---
title: "Breast_cancer"
output: html_document
date: "2022-10-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load library
```{r}
library(tidyverse)
```

# Load the dataset
```{r}
df <- read.csv("METABRIC_RNA_Mutation.csv")
```
## Context 



## Subset the dataset
```{r}
# Keep the first 31 clinical features
df <- df[0:31]
str(df)
```
```{r}
features <- c("patient_id", "age_at_diagnosis", "type_of_breast_surgery", "cancer_type", "chemotherapy", "neoplasm_histologic_grade", "hormone_therapy", "inferred_menopausal_state", "primary_tumor_laterality", "overall_survival_months", "radio_therapy", "tumor_size" , "tumor_stage", "death_from_cancer" )
```

```{r}
df_subset <- df[,features]
str(df_subset)
```

```{r}
summary(df_subset)
```

# Convention for the status

We want to change the columns
0: living or died from other causes
1: died because of the cancer

```{r}
df_subset <- df_subset %>%
  rename(status = "death_from_cancer",
         duration= "overall_survival_months")%>%
  mutate(
    status = ifelse(status == "Died of Disease", 1, 0)
  )
  
  
```

# Overall non parametric survival analysis

```{r}
## Nelson-AAlen estimator
fit_na <- survfit(Surv(duration, status) ~ 1, data = df_subset, type = "fh")
plot(fit_na, 
     main ="Nelson Aalen survival estimator",
     xlab = "Months",
     ylab= "Survival function")
```

```{r}
fit_na
```

> We notice from this survival analysis there are 50% of surivals after 286 months.




## Kaplan-meier with one features and 2 groups: chemiotherapy

```{r}
surv_chem <- survfit(Surv(duration, status ) ~ chemotherapy, data = df_subset)
plot(surv_chem, col= (1:2), main ="Kaplan-Meier", xlab = "Months", ylab = "Survival function")
legend("topright", legend= names(surv_chem$strata),
       col=c("black", "red"), lty=1)
#summary(surv_chem)
```

```{r}
surv_chem
```
> We can observe it seems there is a difference between the 2 groups. indeed, people without any chemiotherapy tend to survive longer  50 % of survivals after 301 months against 147 for people with chemiotherapy.
Let's have a look on the significance of this difference with a LogRank.

## LogRank

Let's define the null hypothesis H0 : the two groups have the same survival distributions.

```{r}
logrank_chem <- survdiff(Surv(duration, status ) ~ chemotherapy, data = df_subset)
logrank_chem
```

> The p-value is << before 0.05. So we can reject the null hypothesis and accept there is a significance difference between the two groups.
Now we will look at an additibve model by includeing with the chemotherapy the raio therapy.

## Kaplan Meir fit with chemotherapy and radio therapy

```{r}
multiple_fit_km <- survfit(Surv(duration, status ) ~ chemotherapy + radio_therapy, data = df_subset)
multiple_fit_km
```

```{r}
plot(multiple_fit_km, col= (1:4), main ="Kaplan-Meier", xlab = "Months", ylab = "Survival functions")
legend("topright", legend= names(multiple_fit_km$strata),
       col=c(1:4), lty=1 )
```
We can observe in this additive model 4 groups:
*   without chemotherapy and without radio therapy: black line
*   without chemotherapy and with radio therapy: red
*   with chemotherapy and without radio therapy: green
*   with chemotherapy and with radio therapy: blue


From this plot we may have a doubt on the significance difference between the group in red (without chemotherapy and with radio therapy) and the one in black ( without chemotherapy and without radio therapy) as there are close together.
Let's do a log rank test.

```{r}
logrank_multiple <- survdiff(Surv(duration, status ) ~ chemotherapy + strata(radio_therapy), data = df_subset)
```

```{r}
logrank_multiple
```

```{r}
survdiff(Surv(duration, status ) ~ chemotherapy[0] + radio_therapy[0], data = df_subset
```












```{r}
